Federated Learning's Blessing:\\ FedAvg has Linear Speedup

### 2020-06-03 source/sec_overparameterized.tex
sec3,4,5 typo

\section{Geometric Convergence of FedAvg in the Overparameterized Setting}
\label{sec:overparameterized}

% In this section, we turn to the special setting of overparameterized
% problems, where a non-negative loss function admits a zero-loss global
% minimizer. Such problems are prevalent especially at the age of deep learning,
% where the number of parameters far exceeds the number of data points,
% and perfect fitting is possible. There have been a line of recent
% works showing that SGD and accelerated methods achieve geometric
% convergence in similar settings thanks to the property of \textit{automatic
% variance reduction}~\cite{ma2017power,moulines2011non,needell2014stochastic,schmidt2013fast,strohmer2009randomized}.
% The natural question is whether such a result also holds in the federated
% learning setting. 
% We show that this is indeed the case and establish
% the geometric (linear) convergence of FedAvg with local SGD updates
% with constant step size for general strongly convex and smooth overparameterized
% problems. In addition, we show that the convergence rate speeds up
% linearly in the number of workers $N$ when $N$ is below some problem-dependent
% threshold, while it decreases with $E$ through $1/E$.

Overparameterization is a prevalent machine learning setting where the
statistical model has much more parameters than the number of training samples
and the existence of parameter choices with zero training loss is
assured~\cite{allen2018convergence,zhang2016understanding}.  Due to the
property of \textit{automatic variance reduction} in overparameterization,  a
line of recent works proved that SGD and accelerated methods achieve geometric
convergence~\cite{ma2017power,moulines2011non,needell2014stochastic,schmidt2013fast,strohmer2009randomized}.
A natural question is whether such a result still holds in the federated
learning setting. 
In this section, we provide the first geometric convergence rate
of FedAvg for the overparameterized strongly convex and smooth problems,
and we show that it preserves the linear speedup at the same time.  
We then sharpen this result in the special case of
quadratic problems.
Inspiring by the recent advances in SGD~\cite{liu2019accelerating,jain2017accelerating}, 
we further propose a novel momentum-based FedAvg algorithm, which enjoys an
improved convergence rate over FedAvg. 
\begin{comment}
we deal with the question of whether FedAvg with momentum-based
local updates can outperform FedAvg with SGD updates. In contrast
to the gradient descent setting, Nesterov and Heavy Ball updates are
known to fail to accelerate over SGD, both in the overparameterized
setting and standard convex setting \cite{liu2018accelerating,kidambi2018insufficiency,liu2018toward,yuan2016influence}.
Thus in general one cannot hope to obtain acceleration results for
the FedAvg algorithm with Nesterov and Heavy Ball updates. On the
hopeful side, \cite{jain2017accelerating,liu2018accelerating} introduced
similar algorithms that make modifications of the Nesterov updates
to correct for its ``over-descent''. For quadratic objectives in
the overparamterized setting, \cite{liu2018accelerating} show their
algorithm achieves acceleration over the geometric convergence of
SGD that recovers the well-known $\mathcal{O}(\exp(-t/\sqrt{\kappa})$
acceleration of Nesterov over GD. In the last part, we introduce a
new accelerated FedAvg algorithm by adapting the MaSS algorithm of
\cite{liu2018accelerating} to the federated learning setting. We
show that it achieves geometric convergence for overparameterized
quadratic problems with rate $\mathcal{O}(\exp(-\frac{NT}{E\sqrt{\kappa_{1}\tilde{\kappa}}}))$,
where $\tilde{\kappa}$ is a ``statistical condition number''\cite{liu2018accelerating,jain2017accelerating}
that satisfies $\tilde{\kappa}\leq\kappa_{1}$. Thus our new FedAvg
algorithm achieves a speedup of factor $\sqrt{\kappa_{1}/\tilde{\kappa}}$
over FedAvg with local SGD updates. 
\end{comment}

>>>>

\section{Geometric Convergence of FedAvg in the Overparameterized Setting}
\label{sec:overparameterized}
% In this section, we turn to the special setting of overparameterized
% problems, where a non-negative loss function admits a zero-loss global
% minimizer. Such problems are prevalent especially at the age of deep learning,
% where the number of parameters far exceeds the number of data points,
% and perfect fitting is possible. There have been a line of recent
% works showing that SGD and accelerated methods achieve geometric
% convergence in similar settings thanks to the property of \textit{automatic
% variance reduction}~\cite{ma2017power,moulines2011non,needell2014stochastic,schmidt2013fast,strohmer2009randomized}.
% The natural question is whether such a result also holds in the federated
% learning setting. 
% We show that this is indeed the case and establish
% the geometric (linear) convergence of FedAvg with local SGD updates
% with constant step size for general strongly convex and smooth overparameterized
% problems. In addition, we show that the convergence rate speeds up
% linearly in the number of workers $N$ when $N$ is below some problem-dependent
% threshold, while it decreases with $E$ through $1/E$. 
Overparameterization is a prevalent machine learning setting where the
statistical model has much more parameters than the number of training samples
and the existence of parameter choices with zero training loss is
assured~\cite{allen2018convergence,zhang2016understanding}.  Due to the
property of \textit{automatic variance reduction} in overparameterization,  a
line of recent works proved that SGD and accelerated methods achieve geometric
convergence~\cite{ma2017power,moulines2011non,needell2014stochastic,schmidt2013fast,strohmer2009randomized}.
A natural question is whether such a result still holds in the federated
learning setting. 
In this section, we provide the first geometric convergence rate
of FedAvg for the overparameterized strongly convex and smooth problems,
and we show that it preserves the linear speedup at the same time.  
We then sharpen this result in the special case of
quadratic problems.
Inspiring by the recent advances in SGD~\cite{liu2019accelerating,jain2017accelerating}, 
we further propose a novel momentum-based FedAvg algorithm, which enjoys an
improved convergence rate over FedAvg. The proof is deferred to Appendix Section~\ref{sec:interpolation}.
\begin{comment}
we deal with the question of whether FedAvg with momentum-based
local updates can outperform FedAvg with SGD updates. In contrast
to the gradient descent setting, Nesterov and Heavy Ball updates are
known to fail to accelerate over SGD, both in the overparameterized
setting and standard convex setting \cite{liu2018accelerating,kidambi2018insufficiency,liu2018toward,yuan2016influence}.
Thus in general one cannot hope to obtain acceleration results for
the FedAvg algorithm with Nesterov and Heavy Ball updates. On the
hopeful side, \cite{jain2017accelerating,liu2018accelerating} introduced
similar algorithms that make modifications of the Nesterov updates
to correct for its ``over-descent''. For quadratic objectives in
the overparamterized setting, \cite{liu2018accelerating} show their
algorithm achieves acceleration over the geometric convergence of
SGD that recovers the well-known $\mathcal{O}(\exp(-t/\sqrt{\kappa})$
acceleration of Nesterov over GD. In the last part, we introduce a
new accelerated FedAvg algorithm by adapting the MaSS algorithm of
\cite{liu2018accelerating} to the federated learning setting. We
show that it achieves geometric convergence for overparameterized
quadratic problems with rate $\mathcal{O}(\exp(-\frac{NT}{E\sqrt{\kappa_{1}\tilde{\kappa}}}))$,
where $\tilde{\kappa}$ is a ``statistical condition number''\cite{liu2018accelerating,jain2017accelerating}
that satisfies $\tilde{\kappa}\leq\kappa_{1}$. Thus our new FedAvg
algorithm achieves a speedup of factor $\sqrt{\kappa_{1}/\tilde{\kappa}}$
over FedAvg with local SGD updates. 
\end{comment}
